A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2019-11-14 paper/main.tex
finalising draft

\begin{abstract}
The Iterated Prisoner's Dilemma has been used for decades as a model of
behavioural interactions. From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of strategies in the game for years. The results of the literature,
however, have been relying on the performance of specific strategies in a finite
number of tournaments, whereas this manuscript evaluates
\numberofstrategies strategies' effectiveness in more than 40000 tournaments.
The top ranked strategies are presented, and moreover, the impact of features on
their success are analysed using machine learning techniques. The analysis
determines that the cooperating ratio of a
strategy in a given tournament compared to the mean and median cooperator is the
most important feature. The conclusions are distinct for different types of
tournaments. A strategy with a theory of mind would aim to be the mean/median
cooperator in standard tournaments, whereas in tournaments with probabilistic
ending it would aim to cooperate 10\% of the times the median cooperator did.
In tournaments with noise and a probability of ending an effective strategy
cooperates 60\% of the times the median/mean cooperator did, and in tourmanents
with only noise or even in a setting that the tournament can be any of the
above a strategy should aim to play 67\% as a median/mean cooperator.
\end{abstract}

@@

Another set of IPD strategies that have received a lot of attention
are the zero-determinant strategies (ZDs)~\cite{Press2012}. By
forcing a linear relationship between the payoffs ZDs can ensure that they will
never receive less than their opponents. The
American Mathematical Society's news section stated that ``the world of game
theory is currently on fire''.
ZDs are indeed a set of mathematically unique strategies
and robust in pairwise interactions, however, their simplicity and extortionate
behaviour have been tested. In~\cite{Harper2017} a tournament containing over
200 strategies, including ZDs, was ran and none of the them
ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based on lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.6\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments. The best performance
    of the strategy has been in standard tournaments where it achieved a $\bar{r}$
    of 0.339.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

@@

In standard tournaments 10 out of the 15 top strategies are introduced
in~\cite{Harper2017}. These are strategies based on finite state automata (FSM),
hidden markov models (HMM), artificial neural networks (ANN), lookup tables
(LookerUp) and stochastic lookup tables (Gambler) that have been trained using
reinforcement learning algorithms (evolutionary and particle swarm algorithms).
They have been trained to perform well against the strategies
in~\cite{axelrodproject} in a standard setting, thus their performance in the
specific setting was anticipated. DoubleCrosser, and Fool Me Once, are
strategies not from the literature but from~\cite{axelrodproject}. DoubleCrosser
is a strategy that makes use of the number of turns because is set to defect on
the last two rounds. The strategy was expected to not perform as well in
tournaments where the number of turns is not specified, but the strategy did not
perform well in tournaments with noise either. Finally, Winner
12~\cite{mathieu2017} and DBS~\cite{Au2006} are both from the the literature.
DBS is strategy specifically designed for noisy environments, however, it ranks
highly only in standard ones.

@@

The top strategies in noisy tournaments are shown in Figure~\ref{fig:noisy_results}. These include deterministic strategies, such
as Tit For 2 Tats~\cite{Axelrod1980b}, Slow Tit For Tats~\cite{axelrodproject}, Hard Tit For 2 Tats~\cite{Stewart2012}
and Cycler CCCCCD, and strategies which decide their actions based on the
cooperations to defections ratio, such as ShortMem~\cite{Carvalho2013}, Grumpy
and $e$~\cite{axelrodproject}. The Retaliate and
Limited Retaliate strategies are implemented in~\cite{axelrodproject} by the
same contributor. They are strategies designed to defect if the opponent has
tricked them more often than \(x\%\) of the times that they have
done the same. Finally, in $4^{\text{th}}$ and $9^{\text{th}}$ place are Hunter
strategies which trying to extort, equivalently, strategies that play cyclically
and defectors.

From Figure~\ref{fig:noisy_results}, it is evident that the normalised rank
distributions in noisy environments are more variant with higher medians
compared to standard tournaments. The distributions are bimodal. This indicates
that although the top ranked strategies mainly performed well, there are several
tournaments that they ranked in the bottom half. To gain a better understanding
of behaviour in noisy tournaments, the \(r\) distribution over the noise
probability \(p_n\) for the top 6 strategies of Figure~\ref{fig:noisy_results},
are given in Figure~\ref{fig:effect_of_noise}.

For \(p_n\) strictly lower than 0.5, Grumpy, Tit For 2 Tats and Slow Tit For Two Tat
have a moderate performing, whereas \(e\), Cycle Hunter and Ricky QLearner are
behaving poorly. At \(p_n=0.5\), all the distributions are bimodal because
with a 0.5 probability of noise any strategies corresponds to a random player.
Interestedly, for a \(p_n\) larger than 0.5 all of the 6 strategies become
successful. Note that a value \(p_n=1\) corresponds to a strategy playing the
exact opposite from what they intended. Thus, the successful strategies in
noisy tournaments presented here, are effective when they are not playing as
themselves. If \(p_n\) was strictly less 0.5 then the top ranked strategies
are different, Table~\ref{table:subset_noise_result}. The top spots are mainly
overtaken by the Meta strategies which are based on teams.

@@

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.30\textwidth}{!}{
    \input{performance_noise_subset_table.tex}}
    \caption{Top performances in noisy tournaments where \(p_n<0.5\).}\label{table:subset_noise_result}
\end{table}

The 15 top ranked strategies in probabilistic ending tournaments include
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014}, which are
strategies based on finite state automata introduced by Daniel and Wendy
Ashlock. These strategies have been evolved using reinforcement learning, however,
there were trained to maximise their payoffs in tournaments with fixed turns
(150 specifically) and not in probabilistic ending ones. In probabilistic ending
tournaments it appears that the top ranks are mostly occupied by defecting
strategies. These include Better and Better, Gradual Killer, Hard Prober (all
from ~\cite{prison}), Bully (Reverse Tit For Tat)~\cite{Nachbar1992} and
Defector. Thus, it's surprisingly that EasyGo and Fool Me Forever which are
strategies that will defect until their opponent defect, then they will cooperate
until the end, ranked $14^{\text{th}}$ and $15^{\text{th}}$. Upon inspection,
it was concluded that they are actually the same strategy. This was not known to
the authors at the time of data collection. Figure~\ref{fig:probend_results}
verifies that their perfomance is the same. Both strategies have repeatedly
ranked highly and there are cases for which they were the winners of the
tournament.

@@

The distributions of \(r\) over \(p_e\) for the top 6 strategies are also given
by Figure~\ref{fig:effect_of_probend}. It is shown that the rank of the
strategies is decreasing as the probability of the game ending increases and at
the point of \(p_e=0.1\) the strategies become dominant. All the top ranked
strategies in this setting are defecting strategies, and what is exhibited is
that they do better as the probability of the game ending in the next turn
increases, which is inline with the Folk Theorem~\cite{Fudenberg2009}. Consider
tournaments where the probability of the game ending was less than 0.1, the top
ranked strategies based on \(\bar{r}\) are given
by~Table~\ref{table:subset_probend_result}. The top ranked strategies include
strategies based on teams, trained strategies, Grudger and Spiteful Tit for Tat.

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.30\textwidth}{!}{
    \input{performance_probend_subset_table.tex}}
    \caption{Top performances in probabilistic ending tournaments with \(p_e < 0.1\)}
    \label{table:subset_probend_result}
\end{table}

In tournaments with both noise and an unspecified number of turns several of
the top ranked strategies are strategies that were highly ranked in noisy
tournaments. This demonstrates that, $e$, $\pi$,
$\phi$ ($\pi$, $\phi$ are based on the same approach as $e$) and the Retaliate
family are robust in noisy environments regardless of the
number of turns. However, strategies from the top ranks in probabilistic
ending tournaments did not rank highly here. The distributions of
\(r\) shown in Figure~\ref{fig:noisy_probend_results} have the largest
median values compared to the top rank strategies of the other tournament
types.

@@

The distributions of \(r\) for the tournament types indicate that
for probabilistic ending and standard tournaments successful strategies do exist.
For these settings, the top 15 strategies have frequently ranked in
the top spots with only a few exceptions. Contrarily, it appears that noise
cause variation in the normalised ranks, and the strategies can always guarantee
a spot in the top ranks.

Up till now, the performances of the \numberofstrategies strategies have been evaluated for
individual tournament types. The data set considered in this work, described in
Section~\ref{section:data_collection}, contains a total of \numberofalltournaments
tournament results. For this part of the manuscript the strategies are ranked
based on the median normalised rank they achieved over the entire data set.
The top 15 strategies are given in Table~\ref{table:overall_results}
and their normalised rank distributions are given in Figure~\ref{fig:overall_results}.

@@

The top ranks include strategies that have been previously mentioned. The
set of Retaliate strategies occupy the top spots followed by BackStabber
and DoubleCrosser. Figure~\ref{fig:overall_results} gives the normalised rank distributions of
these strategies. The distributions of the Retaliate strategies have no
statistical difference. Thus, in an IPD tournament where the type is not
specified, playing as any of the Retaliate strategies will have the result.
DoubleCrosser performed well in standard tournaments and the
strategy is just an extension of BackStabber. It should be noted that these
strategies can be characterised as "cheaters". The source code of the strategies
allows them to known the number of turns in a match (if they are specified).
PSO Gambler and Evolved HMM 5 are
trained strategies introduced in~\cite{Harper2017} and Nice Meta Winner and NMWE
Memory One are strategies based on teams. Grudger is a strategy from Axelrod's
original tournament and Forgetful Fool Me Once is based on the same approach as
Grudger. All the top 15 strategies are fundamentally different. Some are cheaters,
some are complex, others are simple deterministic strategies and strategies based
on teams. The results of \numberofalltournaments tournaments used in this work imply the following:
they is not a single type of strategy which can performance well in any IPD interaction.

@@

This section presented the winning strategies in a series of IPD tournaments. In
standard tournaments the top spots were dominated by complex strategies that had
been trained using reinforcement learning techniques. In noisy environments,
whether the number of turns was fixed or not, the winning strategies were
deterministic strategies designed to defect if the opponent tricked them more
than a current amount of the times that they had tricked their opponent. In
probabilistic ending tournaments most of the winning strategies were defecting
strategies and trained finite state automata, designed by the same authors.
Finally the performance of all \numberofstrategies
strategies over the \numberofalltournaments tournaments in this manuscript was
assessed on \(\bar{r}\). The top ranked strategies were a mixture of behaviours
that did well in standard tournaments and tournaments with noise, as well as a
few strategies based on teams.

The results of this section imply that successful strategies for specific
settings exist for an IPD tournament. The top ranked strategies in both standard
tournaments and tournaments with probabilistic ending, managed to rank in the
top 10\% of the tournament most of the times. Strategies in noisy
environments demonstrated that no strategy can be consistently
successful. Overall, there has been not a single strategy that has shown to perform well in
more than one setting. The aim of the next section is to understand which are
the factors that made these strategies successful, in each setting separately
but also overall.

@@

The performances are clustered in two clusters based on the nornalised rank.
More specifically, there are clustered based on whether the normalised rank was
in the top 5\%, the 25\% and the 50\%. A random forest
approach~\cite{breiman2001} is applied to the data to predict the cluster to
which a strategy's performance has been assigned to. The random forest method
constructs many individual decision trees and the predictions from all trees are
pooled to make the final prediction. The random forest models are trained on a
training set of 70\% of the tournaments results. The accuracy of each model
based on $R^2$ are given by Table~\ref{table:accuracy_random_forest}. The out of
the bag error~\cite{hastie2005} has also been calculated. The models fit well,
and a high value of both the accuracy measure on the test data and the OOB error
indicate that the model is not over fitting.

The performances have also been clustered   based on their normalised rank and
their median score by a \(k-\)means algorithm~\cite{Arthur2007}. The number of
clusters is not deterministically chosen but it is based on the silhouette
coefficients~\cite{Rousseeuw1987}. The chosen cluster for each tournament type,
as well as the accuracy for random forest models, are given in
Table~\ref{table:accuracy_random_forest}.

@@

The importance that the measures of Table~\ref{table:manual_measures} had on the
classification tasks; to which cluster a performance was assigned to, have been
calculated and are given by Figures~\ref{fig:clustering_importance_standard},
\ref{fig:clustering_importance_noise},~\ref{fig:clustering_importance_probend},
\ref{fig:clustering_importance_probend_noise}
and~\ref{fig:clustering_importance_overall}. These show that the classifiers
stochastic, make use of game and make use of length have no significant effect,
and several of the measures that are highted by the importance are inline with the
correlation results. The most important measures based on the random forest
analysis were $C_r / C_{median}$ and $C_r / C_{mean}$.

@@

The effect of both these measures can be further explored. In
Figure~\ref{fig:mean_median_std} the distributions of \(C_ / C_{\text{mean}}\)
and \(C_ / C_{\text{median}}\) are given for the winners. A value of \(C_ /
C_{\text{mean}} = 1\) imply that the cooperating ratio of the winner was the
same as the mean / median cooperating ratio. In standard tournaments, the mean
for both ratios is 1. Therefore, an effective strategy in standard tournaments
was the mean / median cooperator of its respective tournament. In comparison,
Figure \ref{fig:mean_median_noisy} shows the distributions of the measure for
the winners in noisy tournaments. The mean is at 0.67, thereupon the winners
cooperated 67\% of the times the mean / median cooperator did. This analysis is
applied to the rest of the tournaments and the distributions are given by
Figures \ref{fig:mean_median_probend}, \ref{fig:mean_median_probend_noisy} and
\ref{fig:mean_median_overall}. In a tournament with noisy and a probabilistic
ending the winners cooperated 60\%, whereas in settings that the type of the
tournament can vary between the types considered in this work the winners
cooperated 67\% of the times the mean or median cooperator did. Finally, in
probabilistic ending tournament it has already been mentioned that defecting
strategies prevail and this result is once again confirmed in this section.

@@

In this section the effect of several measures, regarding a strategy's behaviour
and the tournament, on it's performance were presented. This was done using
two approaches. Correlation coefficients and a random forest
analysis. The results of these are discussed in the following section.

@@

This manuscript has explored the performance of 194 strategies of the Iterated
Prisoner's Dilemma in a large number of computer tournaments. The results of
the analysis demonstrated that, although for specific tournament type, such as
standard, noisy and probabilistic ending tournaments, dominant strategies exist
there is not a single dominant strategy if the environments
vary. Moreover, a strategy with a theory of mind should aim to adapt its behaviour
based on the mean and median cooperators.

The 194 strategies used in this manuscript have been mainly for the literature,
and they have been accessible due to an open source software called
Axelrod-Python. The software was used to generate a total of
\numberofalltournaments computer tournaments results with different number of
strategies and different participants each time. The data collection was
described in Section~\ref{section:data_collection}. In
Section~\ref{section:top_performances}, the tournaments results were used to
present the top performances.
The data set contained results from four different settings, and these were also
studied individually. In standard tournaments complex strategies trained using
reinforcement learning ranked in the top places. In probabilistic ending
tournaments, several trained strategies based on finite state automata also
performed well. The rest top strategies, in this setting, were defecting
strategies. Finally, in tournaments with noise, the Retaliate set of strategies
demonstrated their robustness in tournaments with probabilistic ending and not.

Section~\ref{section:evaluation_of_performance}, covered an analysis of
performance based on several measures associated with a strategy and with the
environments it was competing. The results of this analysis showed that a
strategy's characteristics such as whether or not it's stochastic, and the information it
used regarding the game had no effect on the strategy's success. The most
important factors have been those that compared the strategy's behaviour to it's
environment. The cooperating ratio of the strategy compared to the mean and
median cooperator was highlighted as the most important feature in the analysis.
More specifically, if a strategy were to enter a tournament with a theory of
mind of it's environment it would choose to be the median cooperator in standard
tournaments, the defector in probabilistic ending tournaments and to cooperate
60\% of the times the median cooperator did in noisy and noisy probabilistic
tournaments. Lastly, if a strategy was aware of the opponents but not of the
setting on the tournament, a strategy would be more likely to be successful if
it were to identify the median cooperator and cooperated 67\% of the times that
they did.

Further data mining could be applied to the data set of this work, available
at~\cite{data}.

>>>>

\begin{abstract}
The Iterated Prisoner's Dilemma has been used for decades as a model of
behavioural interactions. From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of strategies in the game for years. The results of the literature,
however, have been relying on the performance of specific strategies in a finite
number of tournaments, whereas this manuscript evaluates
\numberofstrategies strategies' effectiveness in more than 40000 tournaments.
The top ranked strategies are presented, and moreover, the impact of features on
their success are analysed using machine learning techniques. The analysis
determines that the cooperating ratio of a
strategy in a given tournament compared to the mean and median cooperator is the
most important feature. The conclusions are distinct for different types of
tournaments. For instance a strategy with a theory of mind would aim to be the mean/median
cooperator in standard tournaments, whereas in tournaments with probabilistic
ending it would aim to cooperate 10\% of the times the median cooperator did.
\end{abstract}

@@

Another set of IPD strategies that have received a lot of attention
are the zero-determinant strategies (ZDs)~\cite{Press2012}. By
forcing a linear relationship between the payoffs ZDs can ensure that they will
never receive less than their opponents. The
American Mathematical Society's news section stated that ``the world of game
theory is currently on fire''.
ZDs are indeed a set of mathematically unique strategies
and robust in pairwise interactions, however, their simplicity and extortionate
behaviour have been tested. In~\cite{Harper2017} a tournament containing over
200 strategies, including ZDs, was ran and none of them
ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based on lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.6\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments. The best performance
    of the strategy has been in standard tournaments where it achieved a $\bar{r}$
    of 0.34.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

@@

In standard tournaments 10 out of the 15 top strategies are introduced
in~\cite{Harper2017}. These are strategies based on finite state automata (FSM),
hidden markov models (HMM), artificial neural networks (ANN), lookup tables
(LookerUp) and stochastic lookup tables (Gambler) that have been trained using
reinforcement learning algorithms (evolutionary and particle swarm algorithms).
They have been trained to perform well against the strategies
in~\cite{axelrodproject} in a standard tournament, thus their performance in the
specific setting was anticipated. DoubleCrosser, and Fool Me Once, are
strategies not from the literature but from~\cite{axelrodproject}. DoubleCrosser
is a strategy that makes use of the number of turns because is set to defect on
the last two rounds. The strategy was expected to not perform as well in
tournaments where the number of turns is not specified, but the strategy did not
perform well in tournaments with noise either. Finally, Winner
12~\cite{mathieu2017} and DBS~\cite{Au2006} are both from the the literature.
DBS is strategy specifically designed for noisy environments, however, it ranks
highly only in standard ones.

@@

The top strategies in noisy tournaments are shown in Figure~\ref{fig:noisy_results}. These include deterministic strategies, such
as Tit For 2 Tats~\cite{Axelrod1980b}, Slow Tit For Two Tats~\cite{axelrodproject}, Hard Tit For 2 Tats~\cite{Stewart2012}
and Cycler CCCCCD, and strategies which decide their actions based on the
cooperations to defections ratio, such as ShortMem~\cite{Carvalho2013}, Grumpy
and $e$~\cite{axelrodproject}. Slow Tit For Two Tats is the same strategy as 
Tit For 2 Tats, and at the time of writing this manuscript the
contributors of~\cite{axelrodproject} made a new release where the strategy
has been removed. However, for the purpose of this work the strategy is kept.
The Retaliate and
Limited Retaliate strategies are implemented in~\cite{axelrodproject} by the
same contributor. They are strategies designed to defect if the opponent has
tricked them more often than \(x\%\) of the times that they have
done the same. Finally, in $4^{\text{th}}$ and $9^{\text{th}}$ place are Hunter
strategies which trying to extort, equivalently, strategies that play cyclically
and defectors.

From Figure~\ref{fig:noisy_results}, it is evident that the normalised rank
distributions in noisy environments are more variant with higher medians
compared to standard tournaments. The distributions are bimodal. This indicates
that although the top ranked strategies mainly performed well, there are several
tournaments that they ranked in the bottom half. To gain a better understanding
of the behaviour in noisy tournaments, the \(r\) distributions for the top 6
of Figure~\ref{fig:noisy_results} strategies over the noise probability \(p_n\),
are given in Figure~\ref{fig:effect_of_noise}.

@@

Figure~\ref{fig:effect_of_noise} shows that for \(p_n\) values lower than 0.5
Grumpy, Tit For 2 Tats and Slow Tit For Two Tat perform moderately, and \(e\),
Cycle Hunter and Ricky QLearner perform poorly. At \(p_n=0.5\) all the
distributions become bimodal. This is because with a noise probability of 0.5,
all strategies correspond to a random player. Interestedly, for a \(p_n\) larger
than 0.5 all of the 6 strategies become successful. Note that a value \(p_n=1\)
corresponds to a strategy playing opposite from what it intended to. Thus, it is
demonstrated that the successful strategies is noisy tournaments are sometimes
effective when \(p_n=0.5\) but overall they are very successful whn they are
playing opposite from their original design. If during the data collection a
\(p_n\) strictly less 0.5 was considered then the top ranked strategies would be
different. There are a total of 5661 trials where \(p_n<0.5\) and the top ranked
strategies are given in Table~\ref{table:subset_noise_result}. The median ranks
are lower than before and the top spots are mainly overtaken by Meta strategies
which include NMWE deterministic and NMWE Long Memory. The Meta
strategies~\cite{axelrodproject} create a team of strategies for themselves and
choose to play as a member of their team based on their scores against a given
opponent.

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.30\textwidth}{!}{
    \input{performance_noise_subset_table.tex}}
    \caption{Top performances in 5661 noisy tournaments where \(p_n<0.5\).}\label{table:subset_noise_result}
\end{table}

The 15 top ranked strategies in probabilistic ending tournaments include
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014}, which are
strategies based on finite state automata introduced by Daniel and Wendy
Ashlock. These strategies have been evolved using reinforcement learning, however,
there were trained to maximise their payoffs in tournaments with fixed turns
(150 specifically) and not in probabilistic ending ones. In probabilistic ending
tournaments it appears that the top ranks are mostly occupied by defecting
strategies. These include Better and Better, Gradual Killer, Hard Prober (all
from ~\cite{prison}), Bully (Reverse Tit For Tat)~\cite{Nachbar1992} and
Defector. Thus, it's surprisingly that EasyGo and Fool Me Forever which are
strategies that will defect until their opponent defect, then they will cooperate
until the end, ranked $14^{\text{th}}$ and $15^{\text{th}}$. Upon inspection,
it was found that they are actually the same strategy. This was not known to
the authors at the time of data collection. Figure~\ref{fig:probend_results}
verifies that their performance is the same. Both strategies have repeatedly
ranked highly and there are cases for which they were the winners of the
tournament.

@@

The distributions of \(r\) for the top 6 strategies in probabilistic ending
tournaments over \(p_e\) are given in Figure~\ref{fig:effect_of_probend}.
Figure~\ref{fig:effect_of_probend} shows that the 6 strategies start of with a
high median rank, however, their ranked decreased as the the probability of the
game ending increased and at the point of \(p_e=0.1\) they became the dominant
strategies in their respective tournaments. In essence, what is demonstrated is
that defecting strategies did better when the likelihood of the game ending in
the next turn increased, which is inline with the Folk
Theorem~\cite{Fudenberg2009}. If tournaments where the probability of the game
ending was less than 0.1 were considered then the the top ranked spots are not
dominated by just defecting strategies anymore,
Table~\ref{table:subset_probend_result}. Instead the effective strategies are
now the Meta strategies, trained strategies, Grudger~\cite{axelrodproject} and
Spiteful Tit for Tat~\cite{prison}.

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.30\textwidth}{!}{
    \input{performance_probend_subset_table.tex}}
    \caption{Top performances in 1139 probabilistic ending tournaments with \(p_e < 0.1\)}
    \label{table:subset_probend_result}
\end{table}

In tournaments with both noise and an unspecified number of turns several of the
top ranked strategies are strategies that were highly ranked in noisy
tournaments. However, strategies from the top ranks in probabilistic ending
tournaments did not rank highly here. Other strategies include $\pi$, $\phi$
which are based on the same approach as $e$. The distributions of \(r\) shown in
Figure~\ref{fig:noisy_probend_results} have the largest median values compared
to the top rank strategies of the other tournament types. A subset of noisy
probabilistic ending tournaments has been considered such that \(p_e < 0.1\) and
\(p_n < 0.5\). The top ranked strategies are given in
Table~\ref{table:subset_probend_noise_result} and it is shown that the Meta
strategies which performed well in noisy tournaments with \(p_n < 0.5\), perform
well once again even the number of turns is not specified. Moreover, several
strategies that did well in probabilistic ending tournaments such as Fortress 3,
Fortress 4, Defector and Better and Better are effective here as well.

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.30\textwidth}{!}{
    \input{performance_probend_noise_subset_table.tex}}
    \caption{Top performances in 568 probabilistic ending tournaments with \(p_e < 0.1\) and \(p_n < 0.5\).}
    \label{table:subset_probend_noise_result}
\end{table}

Up till now, the performances of the \numberofstrategies strategies have been evaluated for
individual tournament types. The distributions of \(r\) for the tournament types indicate that
for probabilistic ending and standard tournaments successful strategies do exist.
For these settings, the top 15 strategies have frequently ranked in
the top spots with only a few exceptions. Contrarily, it appears that noise
cause variation in the normalised ranks, and the strategies can always guarantee
a spot in the top ranks.

The data set considered in this work, described in
Section~\ref{section:data_collection}, contains a total of \numberofalltournaments
tournament results. For this part of the manuscript the strategies are ranked
based on the median normalised rank they achieved over the entire data set.
The top 15 strategies are given in Table~\ref{table:overall_results}
and their normalised rank distributions are given in Figure~\ref{fig:overall_results}.

@@

The top ranks include strategies that have been previously mentioned. The
set of Retaliate strategies occupy the top spots followed by BackStabber
and DoubleCrosser. The distributions of the Retaliate strategies have no
statistical difference. Thus, in an IPD tournament where the type is not
specified, playing as any of the Retaliate strategies will have the result.
DoubleCrosser performed well in standard tournaments and the
strategy is just an extension of BackStabber. It should be noted that these
strategies can be characterised as "cheaters". The source code of the strategies
allows them to known the number of turns in a match (if they are specified).
PSO Gambler and Evolved HMM 5 are
trained strategies introduced in~\cite{Harper2017} and Nice Meta Winner and NMWE
Memory One are strategies based on teams. Grudger is a strategy from Axelrod's
original tournament and Forgetful Fool Me Once is based on the same approach as
Grudger. Overall the top 15 strategies are fundamentally different. Some are cheaters,
some are complex, others are simple deterministic strategies and strategies based
on teams. The results of \numberofalltournaments tournaments used in this work imply the following:
they is not a single type of strategy which can performance well in any IPD interaction.

@@

This section presented the winning strategies in a series of IPD tournaments. In
standard tournaments the top spots were dominated by complex strategies that had
been trained using reinforcement learning techniques. In noisy environments,
whether the number of turns was fixed or not, the winning strategies were
deterministic strategies designed to defect if the opponent tricked them more
than a current amount of the times that they had tricked their opponent. However,
if a value of noise strictly less than 0.5 was considered, then the successful
strategies were strategies based on teams. In
probabilistic ending tournaments most of the winning strategies were defecting
strategies and trained finite state automata, designed by the same authors.
These strategies only did better when the probability of the game ending after
each turn was increased.
Finally the performance of all \numberofstrategies
strategies over the \numberofalltournaments tournaments in this manuscript was
assessed on \(\bar{r}\). The top ranked strategies were a mixture of behaviours
that did well in standard tournaments and tournaments with noise, as well as a
few strategies based on teams.

The results of this section imply that successful strategies for specific
settings exist for an IPD tournament. The top ranked strategies in both standard
tournaments and tournaments with probabilistic ending, managed to rank in the
top 10\% of the tournament most of the times.
Strategies in noisy
environments demonstrated that no strategy can be consistently
successful, expected if the value of noise is constrained to less than a half.
Overall, there has been not a single strategy that has shown to perform well in
more than one setting. The aim of the next section is to understand which are
the factors that made these strategies successful, in each setting separately
but also overall.

@@

The performances are clustered based on the normalised rank. More specifically,
they are clustered 3 times into 2 different clusters based on  on whether their
normalised rank was in the top 5\%, 25\% or 50\% respectively. A random forest
approach~\cite{breiman2001} is then applied to each performance to predict the cluster to
which it has been assigned to. The random forest method
constructs many individual decision trees and the predictions from all trees are
pooled to make the final prediction. The random forest models are trained on a
training set of 70\% of the tournaments results. The accuracy of each model
based on $R^2$ are given by Table~\ref{table:accuracy_random_forest}. The out of
the bag error~\cite{hastie2005} has also been calculated. The models fit well,
and a high value of both the accuracy measure on the test data and the OOB error
indicate that the model is not over fitting.

The performances have also been clustered based on their normalised rank and
their median score by a \(k-\)means algorithm~\cite{Arthur2007}. The number of
clusters is not deterministically chosen but it is based on the silhouette
coefficients~\cite{Rousseeuw1987}. The chosen cluster for each tournament type,
as well as the accuracy for random forest models are also given in
Table~\ref{table:accuracy_random_forest}.

@@

The importance that the measures of Table~\ref{table:manual_measures} had on
each classification task; to which cluster a performance was assigned to based
on the normalised rank, and their normalised rank and median score have been
calculated and are given by Figures~\ref{fig:clustering_importance_standard},
\ref{fig:clustering_importance_noise},~\ref{fig:clustering_importance_probend},
\ref{fig:clustering_importance_probend_noise}
and~\ref{fig:clustering_importance_overall}. These show that the classifiers
stochastic, make use of game and make use of length have no significant effect,
and several of the measures that are highted by the importance are inline with
the correlation results. Moreover, the smoothing parameter \(k\) appears to no
have a significant effect either. The most important measures based on the
random forest analysis were $C_{r} / C_{median}$ and $C_r / C_{mean}$.

@@

The effect of both these measures can be further explored. In
Figure~\ref{fig:mean_median_std} the distributions of \(C_r / C_{\text{mean}}\)
and \(C_r / C_{\text{median}}\) are given for the winners in standard tournaments. A value of \(C_r /
C_{\text{mean}} = 1\) imply that the cooperating ratio of the winner was the
same as the mean/median cooperating ratio of the tournament. In standard tournaments, the mean
for both ratios is 1. Therefore, an effective strategy in standard tournaments
was the mean/median cooperator of its respective tournament. In comparison,
Figure~\ref{fig:mean_median_noisy} shows the distributions of the measures for
the winners in noisy tournaments where the mean is at 0.67. Thereupon the winners
cooperated 67\% of the times the mean/median cooperator did. This analysis is
applied to the rest of the tournaments and the distributions are given by
Figures \ref{fig:mean_median_probend}, \ref{fig:mean_median_probend_noisy} and
\ref{fig:mean_median_overall}. In a tournament with noisy and a probabilistic
ending the winners cooperated 60\%, whereas in settings that the type of the
tournament can vary between the types considered in this work the winners
cooperated 67\% of the times the mean or median cooperator did. Finally, in
probabilistic ending tournament it has already been mentioned that defecting
strategies prevail and this result is once again confirmed in this section.

@@

In this section the effect of several measures, regarding a strategy's behaviour
and the tournament in which it participated on its performance were presented.
This was done using two approaches. Correlation coefficients and a random forest
analysis. The results of these are summarised in the following section.

@@

This manuscript has explored the performance of \numberofstrategies strategies of the Iterated
Prisoner's Dilemma in a large number of computer tournaments. The results of
the analysis demonstrated that, although for specific tournament types such as
standard and probabilistic ending tournaments, dominant strategies exist
there is not a single dominant type of strategies if the environments
vary. Moreover, a strategy with a theory of mind should aim to adapt its behaviour
based on the mean and median cooperators.

The \numberofstrategies strategies used in this manuscript have been mainly for
the literature, and they have been accessible due to an open source software
called Axelrod-Python. The software was used to generate a total of
\numberofalltournaments computer tournaments results with different number of
strategies and different participants each time. The data collection was
described in Section~\ref{section:data_collection}. In
Section~\ref{section:top_performances}, the tournaments results were used to
present the top performances. The data set contained results from four different
settings, and these were also studied individually. In standard tournaments
complex strategies trained using reinforcement learning ranked in the top spots.
Some of these strategies ranked again in the top spots in probabilistic
ending tournaments when a \(p_e\) of less 0.1 was considered. In probabilistic
ending tournaments \(p_e\) was designed to vary between 0 and 1. It was demonstrated
that for values larger than 0.1, as stated in the Folk Theorem, defecting strategies
were winning the tournaments because there was a high likelihood of the game
ending in the next turn. In tournaments with noise the median ranks of the top
15 strategies had the highest values and the \(r\) distributions were bimodal.
The top rank strategies were performing both well and bad, and this indicates
that in noisy tournaments there are not strategies that can guarantee winning.
Overall, the top ranked strategies differed from one tournament type to
another and the mechanism behind the winning strategies were all different.
Even strategies designed to do good in one setting did better in others.
On the whole ... (the ipd interactions are unique and there is no winning
strategy) %TODO help from Vince

Section~\ref{section:evaluation_of_performance}, covered an analysis of
performance based on several measures associated with a strategy and with the
environments it was competing. The results of this analysis showed that a
strategy's characteristics such as whether or not it's stochastic, and the information it
used regarding the game had no effect on the strategy's success. The most
important factors have been those that compared the strategy's behaviour to it's
environment. The cooperating ratio of the strategy compared to the mean and
median cooperator was highlighted as the most important feature in the analysis.
More specifically, if a strategy were to enter a tournament with a theory of
mind of its environment it would choose to be the median cooperator in standard
tournaments, to cooperate 10\% in probabilistic ending tournaments and 
60\% in noisy and noisy probabilistic
tournaments of the times the median cooperator did. Lastly, if a strategy was aware of the opponents but not of the
setting on the tournament, a strategy would be more likely to be successful if
it were to identify the median cooperator and cooperated 67\% of the times that
they did.

The data set described in this work contains the largest number of IPD tournaments,
to the authors knowledge, and it available at~\cite{data}. Further data mining
could be applied and provide new insights in the field.
